Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2022-01-22 publications/frontiers/dora_outlier_detection_frontiers.tex
Suggested minor wording updates. (#79)

Despite wide differences in applications, data types, and dimensionality,
the same underlying machine learning algorithms can be employed across 
all of these domains. A challenge for applying them however is that domain
scientists do not always have the programming or machine learning background
to apply the algorithms themselves using existing tools. Given the widespread 
applicability and transferability of OOD methods, the scientific community 
would benefit from a tool that made it easy for them to apply popular outlier
detection algorithms to their science datasets. We created DORA 
(Domain-agnostic Outlier Ranking Algorithms) 
to provide a tool for applying outlier 
detection algorithms to a variety of scientific data sets with minimal coding
required. Users need only to specify details for their data/application 
including the data type, location, and algorithms to run in an experiment
configuration file. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
(DEMUD)~\citep{wagstaff:demud13}, principal component analysis (PCA),
Reed-Xiaoli (RX) detector,  
Local RX, negative sampling~\cite{sipple:neg-sampling20}, and probabilistic 
autoencoder (PAE). 
Each algorithm assigns an outlier 
score to each sample in a given dataset. 
DORA provides results organization and visualization 
modules to help users process the results, including sorting samples by outlier 
score, evaluating outlier recall for a set of known/labeled outliers, clustering 
groups of similar outliers together, and web visualization. 
We demonstrated how DORA
facilitates application, evaluation, and interpretation of outlier detection
methods by performing
experiments for three real-world datasets from Earth science, planetary
science, and astrophysics, as well as one benchmark dataset 
(MNIST/Fashion-MNIST). We found that no single algorithm
performed best across all datasets, underscoring the need for a tool 
that enables comparison of multiple algorithms.

@@

In astrophysics, 
outlier detection methods have been used to identify astrophysical
objects with unique characteristics~\cite{hayat2021self} 
as well as data or modeling 
artifacts in astronomical surveys~\citep{wagstaff:des-anom20}. 
Example outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised}, change~\cite{touati2020anomaly}, 
and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020comparison,
wagstaff:demud13}.
%
These examples show the benefit of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the many available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. DORA aims to fill the need 
for tools that faciliate application, evaluation, and interpretation of outlier
detection methods.

@@

\subsection{Outlier ranking algorithms} 
We implemented 7 unsupervised 
algorithms for scoring and 
ranking samples by outlierness. 
We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may  
perform better for different datasets and use cases. 
We describe each
approach to scoring outliers and the associated methods below.

@@

\subsubsection{Distance}
The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing. RX scores samples using
the Mahalanobis
distance between a sample and a background mean and covariance
~\citep{reed1990adaptive}. The local variant of RX (Local
RX or LRX) can be used for image or raster data and scores each pixel in
an image with respect to a window ``ring'' of pixels surrounding 
it~\citep{molero2013analysis}. 
LRX requires two parameters to define the size of the outer window 
surrounding the pixel and the inner window
around the target pixel to exclude from the background distribution.

\subsubsection{Sparsity}
Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because fewer splits were 
required to isolate the sample.

@@

\subsection{Benchmark: MNIST and Fashion-MNIST}
We used MNIST and F-MNIST to demonstrate DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images 
each from MNIST (outliers) and F-MNIST (inliers). 
%Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.

@@

\section*{Author Contributions}
HRK, UR, and KLW developed the conception of the project. HK oversaw the overall implementation and led the analysis of the Earth science dataset. UR led the analysis of the astrophysics dataset. KLW led the analysis of the planetary science dataset. SL led the development of the DORA software architecture. BD led the analysis of the FMNIST/MNIST dataset. EH supported the analysis of the astrophysics dataset. JL developed the desktop application for results visualization. VR supported the analysis of the Earth science dataset. SK supported the analysis of the planetary science dataset. All authors contributed to the overall system implementation and manuscript writing/editing.

\section*{Funding}
We thank the NASA Center Innovation Fund (\todo{Award number?})
and the NASA Science Mission Directorate 
(under the NASA Cross-Divisional AI Use Case Demonstration grants;
Award \#80NSSC21K0720) for supporting this work.

@@

%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
%%% If using *.tif files convert them to .jpg or .png
%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%

@@

>>>>

Despite wide differences in applications, data types, and dimensionality,
the same underlying machine learning algorithms can be employed across 
all of these domains. A challenge for applying them however is that domain
scientists do not always have the programming or machine learning background
to apply the algorithms themselves using existing tools. Given the widespread 
applicability and transferability of OOD methods, the scientific community 
would benefit from a tool that made it easy for them to apply popular outlier
detection algorithms to their science datasets. We created DORA 
(Domain-agnostic Outlier Ranking Algorithms) 
to provide a tool for applying outlier 
detection algorithms to a variety of scientific data sets with minimal coding
required. Users need only to specify details for their data/application 
including the data type, location, and algorithms to run in an experiment
configuration file. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
(DEMUD)~\citep{wagstaff:demud13}, principal component analysis (PCA),
Reed-Xiaoli (RX) detector~\citep{reed1990adaptive},  
Local RX, negative sampling~\cite{sipple:neg-sampling20}, and probabilistic 
autoencoder (PAE). 
Each algorithm assigns an outlier 
score to each sample in a given dataset. 
DORA provides results organization and visualization 
modules to help users process the results, including sorting samples by outlier 
score, evaluating outlier recall for a set of known/labeled outliers, clustering 
groups of similar outliers together, and web visualization. 
We demonstrated how DORA
facilitates application, evaluation, and interpretation of outlier detection
methods by performing
experiments for three real-world datasets from Earth science, planetary
science, and astrophysics, as well as one benchmark dataset 
(MNIST/Fashion-MNIST). We found that no single algorithm
performed best across all datasets, underscoring the need for a tool 
that enables comparison of multiple algorithms.

@@

In astrophysics, 
outlier detection methods have been used to identify astrophysical
objects with unique characteristics~\cite{hayat2021self} 
as well as data or modeling 
artifacts in astronomical surveys~\citep{wagstaff:des-anom20}. 
Example outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised}, change~\cite{touati2020anomaly}, 
and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020comparison,
wagstaff:demud13}.
%
These examples show the benefit of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the many available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. DORA aims to fill the need 
for tools that facilitate application, evaluation, and interpretation of outlier
detection methods.

@@

\subsection{Outlier ranking algorithms} 
We implemented seven unsupervised 
algorithms for scoring and 
ranking samples by outlierness. 
We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may  
perform better for different datasets and use cases. 
We describe each
approach to scoring outliers and the associated methods below.

@@

\subsubsection{Distance}
The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing. RX scores samples using
the Mahalanobis
distance between a sample and a background mean and covariance~\citep{reed1990adaptive}. The local variant of RX (Local
RX or LRX) can be used for image or raster data and scores each pixel in
an image with respect to a window ``ring'' of pixels surrounding 
it~\citep{molero2013analysis}. 
LRX requires two parameters to define the size of the outer window 
surrounding the pixel and the inner window
around the target pixel to exclude from the background distribution.

\subsubsection{Sparsity}
Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because the number
of random splits required to isolate the sample is small.

@@

\subsection{Benchmark: MNIST and Fashion-MNIST}
We used MNIST and Fashion-MNIST (F-MNIST) to demonstrate DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images 
each from MNIST (outliers) and F-MNIST (inliers). 
%Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.

@@

\section*{Author Contributions}
HRK, UR, and KLW developed the conception of the project. HRK oversaw the overall implementation and led the analysis of the Earth science dataset. UR led the analysis of the astrophysics dataset. KLW led the analysis of the planetary science dataset. SL led the development of the DORA software architecture. BD led the analysis of the FMNIST/MNIST dataset. EH supported the analysis of the astrophysics dataset. JL developed the desktop application for results visualization. VR supported the analysis of the Earth science dataset. SK supported the analysis of the planetary science dataset. All authors contributed to the overall system implementation and manuscript writing/editing.

\section*{Funding}
We thank the NASA Center Innovation Fund 
and the NASA Science Mission Directorate 
(under the NASA Cross-Divisional AI Use Case Demonstration grants;
Award \#80NSSC21K0720) for supporting this work.

@@

%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
%%% Use this if adding the figures directly in the manuscript, if so, please remember to also upload the files when submitting your article
%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
%%% If using *.tif files convert them to .jpg or .png
%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%

@@

% LocalWords:  cls
